### 2017-06-30 paper/lexicographic-programs.tex
removed pre-expectation in favor of ranking transitions re-arranged the section according to this change

We aim to define a function assigning a non-negative vector to each configuration (so called measurable map) such that in each point of computation, the expected value of the function after performing one more computational step is smaller (in lexicographic ordering) than the current one. This property can be formalized using the standard  notion of \emph{pre-expectation}~\cite{xxx}.

@@

\begin{definition}[Pre-Expectation]
	Let $\lem$ 
	be 1-dimensional measurable map for a \PP{} $\program$.
	%Let $\locs$ 
	%be the set of locations of $\pCFG_P$ and $\pvars$ its set of program 
	%variables. 
	The 
	pre-expectation of $\lem$ is a function $\preexp{\lem}\colon \locs\times 
	\Rset^{|\pvars|} \rightarrow \Rset\cup\{\infty\}$ defined as follows:
	\begin{compactitem} %\itemsep1pt \parskip0pt \parsep0pt
		\item 
		if $\loc$ is a probabilistic branching location, then
		$$\preexp{\lem}(\loc,\vec{x}):=\sum_{(\loc,\loc')\in\transitions} 
		Pr_{\loc}\left(\loc,\loc'\right)\cdot
		\lem(\loc',\vec{x});$$
		\item 
		if $\loc$ is a non-deterministic branching location, then
		$$\preexp{\lem}(\loc,\vec{x}):=
		\max_{(\loc,\loc')\in\transitions}\lem(\loc',\vec{x});$$

		%\item 
		%$\mathrm{pre}_\eta(\loc,\mathbf{x}):=\min_{(\loc,id,\ell')\in\transitions}\eta\left(\loc',\mathbf{x}\right)$
		% if $\loc$ is an angelic location;
		\item 
		if $\loc$ is a deterministic location, then for each $\vec{x}$ the value 
		$\preexp{\lem}(\loc,\vec{x})$ is determined as follows: there is exactly one 
		transition
		$\tau=(\loc,\loc')$ such that $\vec{x}\models G(\tau)$. We put $$\preexp{\lem}(\loc,\vec{x}):= \lem(\loc',\vec{x})$$.
		\item 
		If $\loc$ is an assignment location, then there is exactly one transition $\tau=(\loc,\loc')$ outgoing from $\loc$.
		We distinguish 
		three cases, depending on $\updates(\tau)=(j,\up)$ (recall that $\up$ is an update element):
		\begin{compactitem}
			\item If $\up\colon \Rset^{|\pvars|}\rightarrow \Rset$ is a Borel-measurable function, then 
			$$\preexp{\lem}(\loc,\vec{x}):= \lem(\loc',\vec{x}(j\leftarrow \up(\vec{x}))).$$
			\item If $\up$ is a distribution $d$, then $$ \preexp{\lem}(\loc,\vec{x}):= 
			\lem(\loc',\vec{x}(j\leftarrow \E[d])),$$ where $\E[d]$ is the expected value of the 
			distribution $d$.
			\item 
			If $\up$ is a set, then $$ \preexp{\lem}(\loc,\vec{x}):= \sup_{a\in\up}
			\lem(\loc',\vec{x}(j \leftarrow a)).$$
		\end{compactitem}
		%$\mathrm{pre}_\eta(\loc,\mathbf{x}):=\eta(\loc',\expv_{\rvars}\left(f(\vec{x},\mathbf{r})\right))$
		% if $\loc$ is a deterministic location, $(\loc,f,\loc')\in\transitions$ and 
		%$\mathbf{x}\in G(\loc,f,\loc')$, where 
		%$\expv_{\rvars}\left(f(\vec{x},\mathbf{r})\right)$ is the expected value of 
		%$f(\vec{x},\cdot)$. %on $\rvars$ if $\vec{x}$ is treated as a constant vector.
	\end{compactitem}
	%\end{definition}
\end{definition}

Intuitively, a pre-expectation of $\lem$ is a 
function which for each configuration $(\loc,\vec{x})$ returns the maximal
expected value of $\lem$ after one step is made from this configuration, where 
the maximum is taken over all possible non-deterministic choices.
%Note that the only way in which pre-expectation can attain an infinite value 
%is 
%due to non-deterministic assignment. However, if the program in question has 
%\emph{bounded non-determinism}, in the sense that all non-deterministic 
%assignments can only choose the assignment value from a bounded set\footnote{A 
%set $u \subseteq \mathcal{R}$ is bounded if it is contained in some interval 
%of 
%a finite length}, then the pre-expectation of any real-valued measurable map 
%is 
%a real-valued function.

@@

\begin{definition}[Lexicographic Ranking Supermartingale Map]
Let $\eps>0$. An $n$-dimensional \emph{lexicographic $\eps$-ranking supermartingale map} ($\eps$-LexRSM map) for a program $\program$ supported by an invariant map $\inv$ is an $n$-dimensional measurable map $\vec{\lem}=(\lem_1,\dots,\lem_n)$ for $\program$ such that for each configuration $(\loc,\vec{x})$ where $\loc\neq \locterm$ and $\vec{x}\in \inv(\loc)$ the following conditions are satisfied:
 \begin{compactitem}
 	%\item For all $(\loc,\vec{x})\in \locs\times\Rset^{|\pvars|}$ it holds
 	%\item For all  it holds 
 	\item
 	for all $1\leq j \leq n$, $\lem_j(\loc,\vec{x})\geq 0$; and
 	\item 
 	there exists $1\leq j \leq$ n such that
 	\begin{compactitem}
 	\item
 	$\preexp{\lem_j}(\loc,\vec{x}) \leq \lem_j(\loc,\vec{x})-\eps$, and
 	\item
 	for all $1\leq j'<j$ it holds 
 	$\preexp{\lem_{j'}}(\loc,\vec{x}) \leq \lem_{j'}(\loc,\vec{x})$.
 	\end{compactitem}
 \end{compactitem}
If additionally $\lem$ is a linear expression map, then we call it a linear $\eps$-LexRSM map ($\eps$-LinLexRSM).
\end{definition}

@@

\begin{theorem}
\label{thm:lexrsm-programs}
Let $\program$ be a probabilistic program. Assume that there exists an $\eps>0$ 
and an $n$-dimensional $\eps$-LexRSM map $\vec{\lem}=(\lem_1,\dots,\lem_n)$ for 
$\program$ supported 
by some 
invariant map $\inv$. 
Then $\program$ terminates almost surely.
\end{theorem}
\begin{proof}
Let $\sigma$ be any measurable scheduler and $\vecinit\in\vecinitset$ any 
initial variable valuation in $\program$.
We define an $n$-dimensional stochastic process 
$\{\vec{X}_{i}\}_{i=0}^{\infty} $ on the probability space 
$(\OmegaRun,\natfilt,\probm^{\sigma}_{\vecinit})$ such 
that for each 
$i\geq 0$ and $1\leq j 
\leq n$ and each run $\run$ we put $\vecseq{X}{i}{j}(\run) = 
\lem_j(\cfg{\sigma}{i}(\run))$. We claim that $\{\vec{X}_{i}\}_{i=0}^{\infty}$ 
is an $\eps$-LexRSM for the termination time $\ttime$ of $\program$. Clearly 
the process is real-valued, componentwise non-negative, and adapted to the 
canonical filtration of $\natfilt$. It remains to prove that condition (3) in 
Definition~\ref{def:lexrsm} is satisfied. To this end, for each $i\geq 0$ we 
define an almost-sure partition of the set $\{\run\in \OmegaRun\mid 
\ttime(\run) >i\}$ into sets $L^{i}_1,\dots,L^{i}_n$ by putting $L^i_j$ to be 
the set of all runs $\run$ such that $\ttime(\run)>i$ and for $\run$ the index 
$j$ is the smallest one such that $\preexp{\lem_j}(\cfg{\sigma}{i}(\run)) \leq 
\lem_j(\cfg{\sigma}{i}(\run))-\eps$ (which entails 
$\preexp{\lem_{j'}}(\cfg{\sigma}{i}(\run)) =
\lem_{j'}(\cfg{\sigma}{i}(\run))-\eps$ for all $1\leq j'< j$). Due to 
definition of an $\eps$-LexRSM map such a $j$ exists for all 
$\run\in\{\ttime>i\}$ and hence we indeed have a partition. It remains to prove 
that irrespective of the initial choice of $\sigma$ and $\vecinit$ it holds, 
for each $1\leq j 
\leq n$, that $\E^\sigma_{\vecinit}[\vecseq{X}{i+1}{j}\mid 
\natfilt_i]\leq\preexp{\lem_{j}}(\cfg{\sigma}{}) $. This can be achieved by a 
standard~\cite{xxx}, through somewhat technical computation, which we defer to 
\AppendixMaterial.

>>>>

We aim to define a function assigning a non-negative vector to each 
configuration (so called measurable map) such that in each point of 
computation, the expected value of the function after performing one more 
computational step is smaller (in lexicographic ordering) than the current one. 
We formalize this property below.
%This property can be formalized using the standard  notion of 
%\emph{pre-expectation}~\cite{xxx}.

@@

We now formalize the notion of a transition in a pCFG being ranked by a 
measurable map. We first define this notion for transitions that do not go out 
of a probabilistic branching location, as these require a special treatment.

\begin{definition}
\label{def:rank1}
Let $\lem$ be a measurable map, $(\loc,\vec{x})$ be a configuration such that 
$\loc\not\in \locsPB$ and let 
$\tau=(\loc,\loc')$ be a 
transition outgoing from $\loc$. For an $\eps\geq 0$ we say that $\tau$ is 
\emph{$\eps$-ranked by $\lem$ from $(\loc,\vec{x})$} if the following 
conditions are 
satisfied, depending on the type of $\loc$:
\begin{compactitem} %\itemsep1pt \parskip0pt \parsep0pt
%		\item 
%		if $\loc$ is a probabilistic branching location, then
%		$$\preexp{\lem}(\loc,\vec{x}):=\sum_{(\loc,\loc')\in\transitions} 
%		Pr_{\loc}\left(\loc,\loc'\right)\cdot
%		\lem(\loc',\vec{x});$$
		\item 
		if $\loc$ is a deterministic or non-deterministic branching location, 
		then
		$$\lem(\loc',\vec{x}) \leq 
		\lem(\loc,\vec{x})-\eps;$$

		%\item

%%$\mathrm{pre}_\eta(\loc,\mathbf{x}):=\min_{(\loc,id,\ell')\in\transitions}\eta\left(\loc',\mathbf{x}\right)$
		% if $\loc$ is an angelic location;
%		\item 
%		if $\loc$ is a deterministic location, then there is 
%exactly one 
%		transition
%		$\tau=(\loc,\loc')$ such that $\vec{x}\models G(\tau)$. We put 
%$$\preexp{\lem}(\loc,\vec{x}):= \lem(\loc',\vec{x})$$.
		\item 
		If $\loc$ is an assignment location, then we distinguish 
		three cases, depending on $\updates(\tau)=(j,\up)$ (recall that $\up$ 
is an update element):
		\begin{compactitem}
			\item If $\up\colon \Rset^{|\pvars|}\rightarrow \Rset$ is a 
Borel-measurable function, then we require
			$$\lem(\loc',\vec{x}(j\leftarrow 
\up(\vec{x}))) \leq \lem(\loc,\vec{x})-\eps$$
			\item If $\up$ is a distribution $d$, then we require $$ 
			\lem(\loc',\vec{x}(j\leftarrow \E[d])) \leq 
			\lem(\loc,\vec{x})-\eps,$$ 
			where $\E[d]$ is the 
expected value of the 
			distribution $d$.
			\item 
			If $\up$ is a set, then we require $$\sup_{a\in\up}
			\lem(\loc',\vec{x}(j \leftarrow a)) \leq \lem(\loc,\vec{x})-\eps.$$
		\end{compactitem}

%%$\mathrm{pre}_\eta(\loc,\mathbf{x}):=\eta(\loc',\expv_{\rvars}\left(f(\vec{x},\mathbf{r})\right))$
		% if $\loc$ is a deterministic location, 
%$(\loc,f,\loc')\in\transitions$ and 
		%$\mathbf{x}\in G(\loc,f,\loc')$, where 
		%$\expv_{\rvars}\left(f(\vec{x},\mathbf{r})\right)$ is the expected 
%value of 
		%$f(\vec{x},\cdot)$. %on $\rvars$ if $\vec{x}$ is treated as a constant 
%%vector.
	\end{compactitem}
\end{definition}

Since ranking supermartingales are required to decrease on average, for 
individual transitions outgoing from $\locsPB$ it does not make sense to say 
that they are ranked or not. Instead, for each $\loc\in\locsPB$ we consider all 
outgoing transitions together.

\begin{definition}
\label{def:rank2}
Let $\lem$ be a measurable map, and let $(\loc,\vec{x})$ be a configuration with
$\loc\in \locsPB$. For an $\eps\geq 0$ we say that $\loc$ is $\eps$-ranked by 
$\lem$ from $(\loc,\vec{x})$ if 
$$\sum_{(\loc,\loc')\in\transitions} 
		Pr_{\loc}\left(\loc,\loc'\right)\cdot
		\lem(\loc',\vec{x}\leq \lem(\loc,\vec{x})-\eps.$$
\end{definition}

To capture the specific of $\locsPB$, we introduce the 
notion of \emph{generalized transition}.

\begin{definition}
A generalized transition of a pCFG $\pCFG$ is either a transition of $\pCFG$ 
outgoing 
from location not in $\locsPB$ or a location $\loc\in\locsPB$.
\end{definition}

Intuitively, we represent the set of transitions outgoing from $\loc\in\locsPB$ 
by the source location $\loc$.  For generalized transitions 
$\tilde{\tau}=\loc\in\locsPB$ we say that $\tilde{\tau}$ is outgoing from 
$\loc$.

Definitions~\ref{def:rank1} and~\ref{def:rank2} 
define when is a generalized transition $\eps$-ranked by $\lem$ from 
configuration $(\loc,\vec{x})$. We say that a 
generalized transition is \emph{unaffected} by $\lem$ from $(\loc,\vec{x})$ if 
it is $0$-ranked by $\lem$ from $(\loc,\vec{x})$.

%\begin{definition}[Pre-Expectation]
%	Let $\lem$ 
%	be 1-dimensional measurable map for a \PP{} $\program$.
%	%Let $\locs$ 
%	%be the set of locations of $\pCFG_P$ and $\pvars$ its set of program 
%	%variables. 
%	The 
%	pre-expectation of $\lem$ is a function $\preexp{\lem}\colon \locs\times 
%	\Rset^{|\pvars|} \rightarrow \Rset\cup\{\infty\}$ defined as follows:
%	\begin{compactitem} %\itemsep1pt \parskip0pt \parsep0pt
%		\item 
%		if $\loc$ is a probabilistic branching location, then
%		$$\preexp{\lem}(\loc,\vec{x}):=\sum_{(\loc,\loc')\in\transitions} 
%		Pr_{\loc}\left(\loc,\loc'\right)\cdot
%		\lem(\loc',\vec{x});$$
%		\item 
%		if $\loc$ is a non-deterministic branching location, then
%		$$\preexp{\lem}(\loc,\vec{x}):=
%		\max_{(\loc,\loc')\in\transitions}\lem(\loc',\vec{x});$$
%		
%		%\item 
%		
%%%$\mathrm{pre}_\eta(\loc,\mathbf{x}):=\min_{(\loc,id,\ell')\in\transitions}\eta\left(\loc',\mathbf{x}\right)$
%		% if $\loc$ is an angelic location;
%		\item 
%		if $\loc$ is a deterministic location, then for each $\vec{x}$ the 
%value 
%		$\preexp{\lem}(\loc,\vec{x})$ is determined as follows: there is 
%exactly one 
%		transition
%		$\tau=(\loc,\loc')$ such that $\vec{x}\models G(\tau)$. We put 
%$$\preexp{\lem}(\loc,\vec{x}):= \lem(\loc',\vec{x})$$.
%		\item 
%		If $\loc$ is an assignment location, then there is exactly one 
%transition $\tau=(\loc,\loc')$ outgoing from $\loc$.
%		We distinguish 
%		three cases, depending on $\updates(\tau)=(j,\up)$ (recall that $\up$ 
%is an update element):
%		\begin{compactitem}
%			\item If $\up\colon \Rset^{|\pvars|}\rightarrow \Rset$ is a 
%Borel-measurable function, then 
%			$$\preexp{\lem}(\loc,\vec{x}):= \lem(\loc',\vec{x}(j\leftarrow 
%\up(\vec{x}))).$$
%			\item If $\up$ is a distribution $d$, then $$ 
%\preexp{\lem}(\loc,\vec{x}):= 
%			\lem(\loc',\vec{x}(j\leftarrow \E[d])),$$ where $\E[d]$ is the 
%expected value of the 
%			distribution $d$.
%			\item 
%			If $\up$ is a set, then $$ \preexp{\lem}(\loc,\vec{x}):= 
%\sup_{a\in\up}
%			\lem(\loc',\vec{x}(j \leftarrow a)).$$
%		\end{compactitem}
%		
%%%$\mathrm{pre}_\eta(\loc,\mathbf{x}):=\eta(\loc',\expv_{\rvars}\left(f(\vec{x},\mathbf{r})\right))$
%		% if $\loc$ is a deterministic location, 
%%$(\loc,f,\loc')\in\transitions$ and 
%		%$\mathbf{x}\in G(\loc,f,\loc')$, where 
%		%$\expv_{\rvars}\left(f(\vec{x},\mathbf{r})\right)$ is the expected 
%%value of 
%		%$f(\vec{x},\cdot)$. %on $\rvars$ if $\vec{x}$ is treated as a constant 
%%%vector.
%	\end{compactitem}
%	%\end{definition}
%\end{definition}

%Intuitively, a pre-expectation of $\lem$ is a 
%function which for each configuration $(\loc,\vec{x})$ returns the maximal
%expected value of $\lem$ after one step is made from this configuration, where 
%the maximum is taken over all possible non-deterministic choices.
%Note that the only way in which pre-expectation can attain an infinite value 
%is 
%due to non-deterministic assignment. However, if the program in question has 
%\emph{bounded non-determinism}, in the sense that all non-deterministic 
%assignments can only choose the assignment value from a bounded set\footnote{A 
%set $u \subseteq \mathcal{R}$ is bounded if it is contained in some interval 
%of 
%a finite length}, then the pre-expectation of any real-valued measurable map 
%is 
%a real-valued function.

@@

\begin{definition}[Lexicographic Ranking Supermartingale Map]
Let $\eps>0$. An $n$-dimensional \emph{lexicographic $\eps$-ranking supermartingale map} ($\eps$-LexRSM map) for a program $\program$ supported by an invariant map $\inv$ is an $n$-dimensional measurable map $\vec{\lem}=(\lem_1,\dots,\lem_n)$ for $\program$ such that for each configuration $(\loc,\vec{x})$ where $\loc\neq \locterm$ and $\vec{x}\in \inv(\loc)$ the following conditions are satisfied:
 \begin{compactitem}
 	%\item For all $(\loc,\vec{x})\in \locs\times\Rset^{|\pvars|}$ it holds
 	%\item For all  it holds 
 	\item
 	for all $1\leq j \leq n$, $\lem_j(\loc,\vec{x})\geq 0$; and
 	\item 
 	for each generalized transition $\tilde{\tau}$ outgoing from $\loc$ there 
 	exists $1\leq j 
 	\leq$ n such that
 	\begin{compactitem}
 	\item
 	$\tilde{\tau}$ is $\eps$-ranked by $\lem_j$ from $(\loc,\vec{x})$
 	\item
 	for all $1\leq j'<j$ we have that $\tilde{\tau}$ is unaffected by 
 	$\lem_{j'}$ from $(\loc,\vec{x})$.
 	\end{compactitem}
 \end{compactitem}
If additionally $\lem$ is a linear expression map, then we call it a linear $\eps$-LexRSM map ($\eps$-LinLexRSM).
\end{definition}

@@

\begin{theorem}
\label{thm:lexrsm-programs}
Let $\program$ be a probabilistic program. Assume that there exists an $\eps>0$ 
and an $n$-dimensional $\eps$-LexRSM map $\vec{\lem}=(\lem_1,\dots,\lem_n)$ for 
$\program$ supported 
by some 
invariant map $\inv$. 
Then $\program$ terminates almost surely.
\end{theorem}
\begin{proof}
Let $\sigma$ be any measurable scheduler and $\vecinit\in\vecinitset$ any 
initial variable valuation in $\program$.
We define an $n$-dimensional stochastic process 
$\{\vec{X}_{i}\}_{i=0}^{\infty} $ on the probability space 
$(\OmegaRun,\natfilt,\probm^{\sigma}_{\vecinit})$ such 
that for each 
$i\geq 0$ and $1\leq j 
\leq n$ and each run $\run$ we put $\vecseq{X}{i}{j}(\run) = 
\lem_j(\cfg{\sigma}{i}(\run))$. We claim that $\{\vec{X}_{i}\}_{i=0}^{\infty}$ 
is an $\eps$-LexRSM for the termination time $\ttime$ of $\program$. Clearly 
the process is real-valued, componentwise non-negative, and adapted to the 
canonical filtration of $\natfilt$. It remains to prove that condition (3) in 
Definition~\ref{def:lexrsm} is satisfied. To this end, for each $i\geq 0$ we 
define an almost-sure partition of the set $\{\run\in \OmegaRun\mid 
\ttime(\run) >i\}$ into sets $L^{i}_1,\dots,L^{i}_n$ by putting $L^i_j$ to be 
the set of all runs $\run$ such that $\ttime(\run)>i$ and for $\run$ the index 
$j$ is the smallest one such that the $(i+1)$-th transition on $\run$ is ranked 
by $\lem_j$ from $\cfg{\sigma}{i}(\run)$. Due to 
definition of an $\eps$-LexRSM map such a $j$ exists for all 
$\run\in\{\ttime>i\}$ and hence we indeed have a partition. It remains to prove 
that irrespective of the initial choice of $\sigma$ and $\vecinit$ it holds, 
for each $1\leq j 
\leq n$ and $j'<j$, that $\E^\sigma_{\vecinit}[\vecseq{X}{i+1}{j}\mid 
\natfilt_i]\leq X_i[j]-\eps $ on $L_j^i$ and 
$\E^\sigma_{\vecinit}[\vecseq{X}{i+1}{j}\mid 
\natfilt_i]\leq X_i[j] $ on $L_{j'}^i$. This follows easily from 
the definition of $L^{i}_1,\dots,L^{i}_n$ and from the definition of a 
transition being $\eps$-ranked by $\lem_j$.
